---
title: "Final Project"
author: "Hikmat Paudel"
date: "2026-04-20"
output:
  pdf_document: default
  html_document:
        toc: true
        float_toc: true
---

# 1. Project Background

## 1.1. Introduction

Rice (Oryza sativa L.) is a major staple food crop in Nepal and plays a key role in food security and rural livelihoods. Improving agronomic practices is essential to increase productivity and economic returns.

One important management practice is seedling density per hill, which can influence crop growth, yield, and profitability.


## 1.2. Objective

The objective of this study is to evaluate the effect of different seedling densities (1–5 seedlings per hill) on growth, grain yield, and economic performance of spring rice.

## 1.3. Research Question

How does seedling density affect growth, grain yield, and economic return of spring rice?

## 1.4. Experimental Design

The experiment was conducted using a Randomized Complete Block Design (RCBD) with:
- 5 treatments (seedling densities: 1, 2, 3, 4, 5 seedlings per hill)
- 4 replications
Each treatment was randomly assigned within each replication.

## 1.5. Dataset Description

The data-set contains 20 observations collected from a field experiment. The variables include:

- rep: replication/block  
- trt: treatment (seedling density)  
- ph: plant height (cm)  
- nt: number of tillers per square meter  
- et: effective tillers per square meter  
- pl: panicle length (cm)  
- fg: filled grains per panicle  
- tgw: thousand grain weight  
- sp: sterility percentage  
- gy: grain yield (tons/ha)  
- sy: straw yield (tons/ha)  
- hi: harvest index  
- gr: gross return (Nepali Rupees per ha)  
- nr: net return (Nepali Rupees per ha)  
- bc: benefit-cost ratio  

## 1.6. Data Import and Preparation

```{r}
library(readr)
library(dplyr)
library(tidyverse)

data <- read.csv("ProjectData.csv")

data <- data %>%
  mutate(rep = as.factor(rep),trt = as.factor(trt))

str(data)
summary(data)
```

# 2. Data Visualization
Data visualization was carried out using ggplot2 to explore treatment effects and relationships among agronomic and economic variables. Figures were prepared with consistent themes, clear labeling, and minimal visual clutter.

## 2.1. Effect of Seedling Density on Grain Yield

```{r}
ggplot(data, aes(x = trt, y = gy)) +
  geom_boxplot(fill = "grey85", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  labs(x = "Seedling density per hill", 
       y = "Grain yield (t/ha)") +
  theme_classic(base_size = 14)
```

This figure shows the distribution of grain yield across seedling density treatments. It highlights both variability and central tendency within each treatment.

## 2.2. Treatment Mean Response (Yield Trend)

```{r}
ggplot(data, aes(x = trt, y = gy)) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "red") +
  stat_summary(fun = mean, geom = "line", aes(group = 1)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.15) +
  labs(x = "Seedling density per hill", 
       y = "Mean grain yield (t/ha)") +
  theme_classic(base_size = 14)
```

This figure presents mean grain yield across treatments with standard error bars, allowing comparison of productivity trends across seedling densities.

## 2.3. Plant Height and Grain Yield Relationship

```{r}
ggplot(data, aes(x = ph, y = gy)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(x = "Plant height (cm)",
       y = "Grain yield (t/ha)") +
  theme_classic(base_size = 14)
```

This figure shows the relationship between plant height and grain yield, indicating how vegetative growth contributes to yield formation.

## 2.4. Tillering Effect on Grain Yield (Multi-variable plot)

```{r}
ggplot(data, aes(x = nt, y = gy, color = ph)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_color_gradient(low = "blue", high = "red") +
  labs(x = "Number of tillers per square meter",
       y = "Grain yield (t/ha)",
       color = "Plant height") +
  theme_classic(base_size = 14)
```

This figure illustrates how tiller density and plant height jointly influence grain yield, highlighting multiple yield-contributing traits.

## 2.5. Yield and Economic Return Relationship

```{r}
ggplot(data, aes(x = gy, y = nr)) +
  geom_point(aes(size = bc), alpha = 0.7, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 1) +
  labs(x = "Grain yield (t/ha)",
       y = "Net return (NPR/ha)",
       size = "Benefit-cost ratio") +
  theme_classic(base_size = 14)
```

This figure shows the economic impact of grain yield, indicating that higher yield is associated with increased net returns and profitability.

## 2.6. Yield Stability Across Replications (RCBD Structure)

```{r}
ggplot(data, aes(x = trt, y = gy, group = rep)) +
  geom_line(alpha = 0.5, color = "grey40") +
  geom_point(aes(color = rep), size = 3) +
  labs(x = "Seedling density per hill",
       y = "Grain yield (t/ha)") +
  theme_classic(base_size = 14)
```

This figure illustrates variation in treatment response across replications, reflecting the structure of the randomized complete block design.

## 2.7. Economic Efficiency (Benefit–Cost Ratio)

```{r}
ggplot(data, aes(x = trt, y = bc)) +
  geom_boxplot(fill = "grey85", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(x = "Seedling density per hill", 
       y = "Benefit–cost ratio") +
  theme_classic(base_size = 14)
```

This figure compares economic efficiency across seedling density treatments, showing differences in profitability among management practices.

# 3. Statistical Analysis

## 3.1. Effect of Seedling Density on Grain Yield

```{r}
model_gy <- lm(gy ~ trt, data = data)
summary(model_gy)
anova(model_gy)
```

This model evaluates whether seedling density significantly affects grain yield. The ANOVA tests whether mean yield differs across treatments.

## 3.2. Mean Separation (Post-hoc Test)

```{r}
library(emmeans)
emmeans_gy <- emmeans(model_gy, ~ trt)
pairs(emmeans_gy)
```

Post-hoc comparisons were used to identify which treatment levels differ significantly in grain yield.

## 3.3. Relationship Between Plant Height and Grain Yield

```{r}
model_ph <- lm(gy ~ ph, data = data)
summary(model_ph)
```

This regression model examines the relationship between plant height and grain yield.

## 3.4. Multiple Regression Model

```{r}
model_multi <- lm(gy ~ ph + nt + et + fg + tgw, data = data)
summary(model_multi)
```

This model evaluates the combined effects of key agronomic traits on grain yield.

## 3.5. RCBD Model (Including Replication)

```{r}
model_rcbd <- lm(gy ~ trt + rep, data = data)
summary(model_rcbd)
anova(model_rcbd)
```

This model accounts for the randomized complete block design by including replication as a factor.

# 4. Model Diagnostics

```{r}
par(mfrow = c(2,2))
plot(model_gy)
```

Diagnostic plots were used to evaluate model assumptions such as normality and homogeneity of variance.

# 5. Reproducible and Efficient Coding

## 5.1. Data Wrangling

```{r}
data_summary <- data %>%
  group_by(trt) %>%
  summarise(
    mean_gy = mean(gy),
    se_gy = sd(gy)/sqrt(n()),
    mean_nr = mean(nr),
    mean_bc = mean(bc))
data_summary
```

## 5.2. Function for Linear Models

```{r}
run_model <- function(response) {
  formula <- as.formula(paste(response, "~ trt"))
  model <- lm(formula, data = data)
  summary(model)
}

run_model("gy")
run_model("ph")
run_model("nt")
```

## 5.3. Iteration (Loop)

```{r}
traits <- c("gy", "ph", "nt", "et", "fg")

results <- lapply(traits, function(x) {
  model <- lm(as.formula(paste(x, "~ trt")), data = data)
  anova(model)
})

names(results) <- traits
results
```

## 5.4. Data Check

```{r}
if(any(is.na(data))) {
  print("Missing values detected")
} else {
  print("No missing values in dataset")
}
```

# 6. Key Findings

- Grain yield varied significantly across seedling density treatments, with intermediate   densities (2–3 seedlings per hill) generally producing higher yields.
- Very low or very high seedling densities resulted in comparatively lower yield.
- Plant height showed a positive association with grain yield, indicating the importance of vegetative growth.
- Tiller number and effective tillers contributed to yield variation in the multiple regression model.
- Grain yield was positively related to net return, indicating higher productivity leads to better profitability.
- Inclusion of replication in the RCBD model improved the precision of treatment comparisons.

# 7. Reproducibility Statement

This analysis was conducted using R and R Markdown. All steps, including data import, visualization, and statistical analysis, are fully reproducible.


